
---
title: "Analysis of scRNA-seq data with Seurat"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r,child="assets/header-lab.Rmd"}
```

```{r, include=FALSE}
knitr::opts_chunk$set(chunk.title = TRUE, engine.opts = list(bash = "-l"), fig.align = "center", eval=FALSE)
```
# Background: Seurat and Single-Cell Correlation

## What is Seurat?

Seurat is a comprehensive R package designed for the analysis, exploration, and visualization of single-cell RNA sequencing (scRNA-seq) data. It provides tools for importing UMI count matrices, performing quality control, normalizing single-cell data, identifying highly variable genes, reducing dimensionality (PCA, UMAP), clustering cells, identifying marker genes, and integrating data from multiple samples or conditions.

Seurat organizes data into a structured Seurat object that contains raw counts, normalized values, dimensional reductions, clustering results, graphs, and metadata. This layered structure enables a reproducible workflow from raw data to biological interpretation.

## What does correlation mean in single-cell data?

Single-cell correlation can be understood in two ways:

### Cell–cell correlation
Each cell is represented by its gene expression profile. Cells that are transcriptionally similar have highly correlated profiles.

Seurat uses these correlations when constructing:
- PCA embeddings  
- k-nearest neighbor (KNN) graphs  
- shared nearest neighbor (SNN) clustering  
- UMAP embeddings  

Clusters of cells represent groups with high similarity, reflecting shared biological identity or state.

### Gene–gene correlation
Genes can also correlate across cells. Correlated genes are often:
- Co-regulated  
- Part of the same pathway  
- Marker genes for the same population  

In Seurat, gene–gene correlation underlies:
- Identification of highly variable genes  
- Cluster-specific marker detection  
- Co-expression visualization (FeaturePlot, DotPlot)

### Why correlations behave differently in single-cell data

Single-cell UMI matrices are sparse and zero-inflated due to technical dropout, meaning many entries are zero even when genes are expressed. As a result, raw correlation matrices are noisy. Seurat mitigates this by conducting downstream analysis in PCA space, which reduces noise and preserves meaningful variation, allowing more robust detection of structure in the data.

Together, these concepts form the foundation for clustering, visualization, and annotation in Seurat.

# Introduction

This tutorial walks through the initial stages of a **single-cell RNA sequencing (scRNA-seq)** workflow using the **Seurat v5** R package. Unlike bulk RNA-seq, which measures average expression across large populations of cells, scRNA-seq provides expression profiles for **individual cells**, enabling fine‑grained analysis of heterogeneity, rare cell types, and dynamic cell states.

We use data from **Nanduri et al. (2022)** (*Nature Communications*), examining the role of HMGN proteins in adipocyte biology by comparing **wild-type (WT)** and **double knockout (DKO)** cells across two time points.

This tutorial covers:

1. Importing 10X Genomics data
2. Creating individual Seurat objects
3. Merging objects
4. Accessing metadata
5. Adding experimental annotations
6. Saving the merged Seurat object


# Data

The dataset includes eight samples:

Sample | Species | Condition | Time Point | Replicate | Status
------ | -------- | ---------- | ----------- | --------- | -------
W10 | Mouse | WT | 0 | 1 | Preadipocytes
W20 | Mouse | WT | 0 | 2 | Preadipocytes
D10 | Mouse | DKO | 0 | 1 | Preadipocytes
D20 | Mouse | DKO | 0 | 2 | Preadipocytes
W16 | Mouse | WT | 6 | 1 | Adipocytes
W26 | Mouse | WT | 6 | 2 | Adipocytes
D16 | Mouse | DKO | 6 | 1 | Adipocytes
D26 | Mouse | DKO | 6 | 2 | Adipocytes

Each sample includes filtered UMI count matrices (`filtered_feature_bc_matrix.h5`) generated by 10X Cell Ranger.

# Installing Seurat

Run this in your own environment:

```{r eval=FALSE}
install.packages("Seurat", repos="https://cloud.r-project.org")
```

Load required packages:

```{r eval=FALSE}
library(Seurat)
library(tidyverse)
```

# Importing Data

List available `.h5` files:

```{r eval=FALSE}
files <- list.files(path="../data/", recursive=TRUE, pattern="*.h5")
files
```

Load each file using Seurat's `Read10X_h5()`:

```{r eval=FALSE}
h5_read <- lapply(paste0("../data/", files), Read10X_h5)
names(h5_read) <- c("D10","D16","D20","D26","W10","W16","W20","W26")
```

Inspect a preview:

```{r eval=FALSE}
h5_read[["W10"]][1:5,1:5]
```

# Creating Seurat Objects

Create objects for each dataset:

```{r eval=FALSE}
adp <- mapply(
  CreateSeuratObject,
  counts = h5_read,
  project = names(h5_read),
  MoreArgs = list(min.cells = 3, min.features = 200)
)
```

Merge into a single object:

```{r eval=FALSE}
adp <- merge(adp[[1]], y = adp[2:length(adp)], add.cell.ids = names(adp), project="Adipose")
```

# Inspecting the Seurat Object

Check structure:

```{r eval=FALSE}
glimpse(adp)
```

Preview metadata:

```{r eval=FALSE}
head(adp@meta.data)
```

Accessing metadata columns:

```{r eval=FALSE}
head(adp$orig.ident)
```

# Adding Metadata

Condition (WT / DKO):

```{r eval=FALSE}
adp$condition <- ifelse(str_detect(adp$orig.ident, "^W"), "WT", "DKO")
```

Time point:

```{r eval=FALSE}
adp$time_point <- ifelse(str_detect(adp$orig.ident, "0"), "Day 0", "Day 6")
```

Combined label:

```{r eval=FALSE}
adp$cond_tp <- paste(adp$condition, adp$time_point)
```

# Checking Cells and Features

```{r eval=FALSE}
head(Cells(adp))
head(Features(adp))
dim(adp)
```

# Saving the Object

```{r eval=FALSE}
saveRDS(adp, "../outputs/merged_Seurat_adp.rds")
```

# Session information

```{r eval=FALSE}
sessionInfo()
```

```{r,child="assets/footer-lab.Rmd", eval=FALSE}
```
